\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-18 paper/main.tex
fixed GRRM citation, updated Fig 2 caption, \citep -> \cite, add non-breaking spaces before \cite and \ref

The \texttt{davos} package is named after Davos Seaworth, a smuggler
referred to as ``the Onion Knight" from the series \textit{A Song of
  Ice and Fire} by George R. R. Martin~\citep{Mart12}.  The
\texttt{smuggle} keyword-like operator implemented in \texttt{davos}
is a play on Python's \texttt{import} keyword: whereas importing can
bring a package into the Python workspace within the existing rules
and frameworks provided by the Python language, ``smuggling'' provides
an alternative that expands the scope and reach of ``importing.''
Like the character Davos Seaworth (who became famous for smuggling
onions through a blockade on his homeland), we use ``onion'' comments
to precisely control how packages are smuggled into the Python
workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The \texttt{davos} package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring \texttt{davos}'s behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual \texttt{.py} files are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a file that imports objects defined or specified at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

IPython preprocesses all executed code as plain text before it is sent
to the Python compiler, in order to handle special constructs like
\texttt{\%magic} and \texttt{!shell} commands. \texttt{davos} uses
this process to transform \texttt{smuggle} statements into
syntactically valid Python code. The \texttt{davos} parser uses a
regular expression to match lines of code containing \texttt{smuggle}
statements (and, optionally, onion comments), extract relevant
information from their text, and replace them with equivalent calls to
the \texttt{smuggle()} function. For example, if a user runs a
notebook cell containing
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet6}
\end{center}
the code that is actually executed by the Python interpreter would be
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet7}
\end{center}
The call to the \texttt{smuggle()} function carries out
\texttt{davos}'s central logic by determining whether the smuggled
package must be installed, carrying out the installation if necessary,
and subsequently loading it into the namespace. This process is
outlined in Figure \ref{fig:flow-chart}. Because the
\texttt{smuggle()} function is defined in the notebook namespace, it
is also possible (though never necessary) to call it
directly. Deactivating \texttt{davos} will delete the name
``\texttt{smuggle}'' from the namespace, unless its value has been
overwritten and no longer refers to the \texttt{smuggle()}
function. It will also deregister the \texttt{davos} parser from the
set of input transformers run when each notebook cell is
executed. While the overhead added by the \texttt{davos} parser is
minimal, this may be useful, for example, when optimizing or precisely
profiling code.

@@

Next, we load in the reformatted dataset (line 32) and pre-trained
model (line 34) that we wish to use in our analysis.  In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pretrained models in succession.  First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts.  Second, the
word counts are passed to a topic model~\citep{BleiEtal03} using a
pretrained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends.  Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions.  For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3).  In that version of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class is defined in
\texttt{sklearn.decomposition.\_lda}.  However, in version
0.22.0, that module was renamed to \texttt{\_online\_lda}, and in
versions 0.22.1 and higher, the name was reverted back to
\texttt{\_lda}.

@@

A second more subtle impact of our work is in providing a
proof-of-concept of how the ability to add new ``keyword-like''
operators to the Python language could be specifically useful to
researchers. With \texttt{davos}, we accomplish this by leveraging
IPython notebooks' internal code parsing and execution machinery. We
note that, while other popular packages similarly use these mechanisms
to providing notebook-specific functionality (e.g.,
\cite{Hunt07,HeusEtal18a}), this approach also has the potential to be
exploited for more nefarious purposes. For example, a malicious user
could design a Python package that, when imported, substantially
changes the notebook's functionality by adding new \textit{unexpected}
keyword-like objects (e.g., based around common typos). We also note
that this implementation approach means \texttt{davos}'s functionality
is currently restricted to IPython notebook environments. However,
there have been early-stage discussions of providing this sort of
syntactic customizability as a core feature of the Python language,
including a draft proposal \cite{Shan20}. In addition to enabling
\texttt{davos} to be extended for use outside of notebooks, this could
lead to exciting new tools that, like \texttt{davos}, extend the
Python language in useful and more secure ways.

>>>>

The \texttt{davos} package is named after Davos Seaworth, a smuggler
referred to as ``the Onion Knight" from the series \textit{A Song of
  Ice and Fire} by George R. R. Martin~\cite{Mart98}.  The
\texttt{smuggle} keyword-like operator implemented in \texttt{davos}
is a play on Python's \texttt{import} keyword: whereas importing can
bring a package into the Python workspace within the existing rules
and frameworks provided by the Python language, ``smuggling'' provides
an alternative that expands the scope and reach of ``importing.''
Like the character Davos Seaworth (who became famous for smuggling
onions through a blockade on his homeland), we use ``onion'' comments
to precisely control how packages are smuggled into the Python
workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The \texttt{davos} package (orange)
  comprises two interdependent subpackages (purple).  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring \texttt{davos}'s behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

IPython preprocesses all executed code as plain text before it is sent
to the Python compiler, in order to handle special constructs like
\texttt{\%magic} and \texttt{!shell} commands. \texttt{davos} uses
this process to transform \texttt{smuggle} statements into
syntactically valid Python code. The \texttt{davos} parser uses a
regular expression to match lines of code containing \texttt{smuggle}
statements (and, optionally, onion comments), extract relevant
information from their text, and replace them with equivalent calls to
the \texttt{smuggle()} function. For example, if a user runs a
notebook cell containing
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet6}
\end{center}
the code that is actually executed by the Python interpreter would be
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet7}
\end{center}
The call to the \texttt{smuggle()} function carries out
\texttt{davos}'s central logic by determining whether the smuggled
package must be installed, carrying out the installation if necessary,
and subsequently loading it into the namespace. This process is
outlined in Figure~\ref{fig:flow-chart}. Because the
\texttt{smuggle()} function is defined in the notebook namespace, it
is also possible (though never necessary) to call it
directly. Deactivating \texttt{davos} will delete the name
``\texttt{smuggle}'' from the namespace, unless its value has been
overwritten and no longer refers to the \texttt{smuggle()}
function. It will also deregister the \texttt{davos} parser from the
set of input transformers run when each notebook cell is
executed. While the overhead added by the \texttt{davos} parser is
minimal, this may be useful, for example, when optimizing or precisely
profiling code.

@@

Next, we load in the reformatted dataset (line 32) and pre-trained
model (line 34) that we wish to use in our analysis.  In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pretrained models in succession.  First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts.  Second, the
word counts are passed to a topic model~\cite{BleiEtal03} using a
pretrained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends.  Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions.  For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3).  In that version of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class is defined in
\texttt{sklearn.decomposition.\_lda}.  However, in version
0.22.0, that module was renamed to \texttt{\_online\_lda}, and in
versions 0.22.1 and higher, the name was reverted back to
\texttt{\_lda}.

@@

A second more subtle impact of our work is in providing a
proof-of-concept of how the ability to add new ``keyword-like''
operators to the Python language could be specifically useful to
researchers. With \texttt{davos}, we accomplish this by leveraging
IPython notebooks' internal code parsing and execution machinery. We
note that, while other popular packages similarly use these mechanisms
to providing notebook-specific functionality (e.g.,
\cite{Hunt07,HeusEtal18a}), this approach also has the potential to be
exploited for more nefarious purposes. For example, a malicious user
could design a Python package that, when imported, substantially
changes the notebook's functionality by adding new \textit{unexpected}
keyword-like objects (e.g., based around common typos). We also note
that this implementation approach means \texttt{davos}'s functionality
is currently restricted to IPython notebook environments. However,
there have been early-stage discussions of providing this sort of
syntactic customizability as a core feature of the Python language,
including a draft proposal~\cite{Shan20}. In addition to enabling
\texttt{davos} to be extended for use outside of notebooks, this could
lead to exciting new tools that, like \texttt{davos}, extend the
Python language in useful and more secure ways.
